Pushdown Control-Flow Analysis via Refunctionalization

### 2018-02-23 paper.tex
update

theoretical and practical

\begin{itemize}
  \item We fill the gap between small-step AAM and big-step ADI 
    by developing a series of systematic transformations from 
    small-step AAM to big-step ADI, and also the reversed 
    direction is also possible.

  \item We present an abstract interpreter written in continuation-passing
    style that is obtained from refunctionalization of a small-step AAM.
    This abstract interpreter naturally has pushdown control-flow property, 
    and we show how precise call/return match is obtained by 
    representing evaluation context as a continuation function, 
    as well as how it is lost by representing continuation as a 
    data type in the defining language.
    As a form of abstract interpreter, We also analyze its termination,
    soundness and complexity property.

  \item We show that the correspondence not only exists between 
    concrete abstract machines and evaluator, and also exists 
    for abstracting abstract machines and abstract evaluators. 
    Galois connection?
\end{itemize}

@@

In this section, we frist describe the target language we will
analyze and a concrete CESK machine as operational semantics for it.
Then we review several recent works in control flow analysis for
higher-order languages:
1) abstracting abstract machines (AAM) as an abstract interpretation
counterpart of CESK machines,
2) definitional abstract interpreter written in a monadically-parameterized
style, and
3) precise call/return match problem in control-flow analysis and its existing solutions.
The abstracting abstract machines form the basis of our transformations, and
monadic abstract interpreter is the end point. In the meanwhile, the precise
call/return match is obtained during transformation.

@@

To begin, we now present the syntax of a call-by-value $\lambda$-calculus language in
administrative normal form (ANF) \cite{flanagan1993essence}.

@@

The abstract syntax tree can be straightforwardly described in Scala as follows:

@@

@@

It is worth noting that the continuation class \texttt{Kont} is defined as a list of frames.
The frame class \texttt{Frame} stores the information of call-site, i.e., the information that
used to resume the interrupted computation.
A \texttt{Frame} constitutes of a variable name \texttt{x} to be bind later, a control expression
that resumes to, and its environment.

Before go into describing how the machine evaluates expressions, we firstly define several helper functions.
As we mentioned in section \ref{anfsyntax}, the atomic expressions are either a variable, or
a literal \texttt{lambda} term. So the helper function \texttt{atomicEval} handles these two
cases and evaluates atomic expressions to closures in a straightforward way.
The \texttt{alloc} function generates a fresh address, and always allocates an unique integer under the domain
of store.
The \texttt{isAtomic} function is used as a predicate of whether the expression is atomic.

@@

To run the program,

@@

\subsection{Abstracting Abstract Machine}

\subsection{Monadic Abstract Interpreter}

@@

@@

\subsubsection{Computable Solutions}
P4F ANF, not works on direct style,PDCFA,AAC

@@

>>>>

\begin{itemize}
\item The main contribution of this paper is we fill the gap
  between small-step AAM and big-step ADI by developing a series of
  systematic transformations. The correspondence between abstract
  machines and interpreters exists not only in concrete semantics artifact,
  but also in abstract semantics artifact.
  Galois connection?

\item We present an abstract interpreter obtained from refunctionalizing
  an small-step abstracting abstract machine. 
  This abstract interpreter naturally has pushdown control-flow property.
  We also analyze its termination, soundness and complexity.

\item Further, we show that defunctionalization and refunctionalization
  of abstract interpreters play important roles for call stack of analyzed language.
  Given this insight, we reveal that why small-step AAM has no pushdown
  control-flow property (if without extra effort), as well as explain
  that why ADI naturally extends pushdown control-flow property from its
  defining language.

\end{itemize}

@@

In this section, we frist describe the target language we will
analyze and a concrete CESK machine as operational semantics for it.
Then we review some recent works in control-flow analysis for
higher-order languages:
1) abstracting abstract machines (AAM) as an abstract interpretation
counterpart of CESK machines, and
2) the precise call/return match problem in control-flow analysis and its existing solutions.
We will show in section 3 that the abstracting abstract machines form the basis of our transformations,
and call/return match is obtained during transformation.

@@

To begin, we now present the concrete syntax of a call-by-value $\lambda$-calculus language in
administrative normal form (ANF) \cite{flanagan1993essence}.

@@

The abstract syntax can be straightforwardly described in Scala as follows:

@@

\subsubsection{Machine Components}

@@

It is worth noting that the continuation class \texttt{Kont} is defined as a list of frames.
The frame class \texttt{Frame} stores the information of call-site, i.e., the information that
can be used to resume the interrupted computation.
A \texttt{Frame} constitutes of a variable name \texttt{x} to be bind later, a control expression
that resumes to, and its environment.

\subsubsection{Single-step Evaluation}
Before go into describing how the machine evaluates expressions, we firstly define several helper functions.
As we mentioned in section \ref{anfsyntax}, the atomic expressions are either a variable, or
a literal \texttt{lambda} term. So the helper function \texttt{atomicEval} handles these two
cases and evaluates atomic expressions to closures in a straightforward way.
The \texttt{alloc} function generates a fresh address, and always allocates an unique integer under the domain
of store.
The \texttt{isAtomic} function is used as a predicate of whether the expression is atomic.

@@

\subsubsection{Valuation}

To run the program, we use function \texttt{inject} to construct an initial machine
state given an expression. Then the \texttt{drive} function is used to evaluate
to a final state by iteratively apply \texttt{step} on current state until we reach a state
that the control is an atomic expression and the continuation stack is empty.
And of course, we can extract the value from the final state at last.

@@

\subsection{Abstracting Abstract Machine} \label{aam}

store allocated continuation
binding store, continuation store
store maps from addresses to a set of values
store update is a join
the step function is non-deterministic

\begin{verbatim}
case class Store[K,V](map: Map[K, Set[V]]) {
  def apply(addr: K): List[V] = map(addr).toList
  def update(addr: K, d: Set[V]): Store[K,V] = {
    val oldd = map.getOrElse(addr, Set())
    Store[K, V](map ++ Map(addr -> (d ++ oldd)))
  }
  def update(addr: K, sd: V): Store[K,V] = update(addr, Set(sd))
}

type Time = List[Expr]

case class BAddr(x: String, time: Time)
type Env = Map[String, BAddr]

abstract class Storable 
case class Clos(v: Lam, env: Env) extends Storable
type BStore = Store[BAddr, Storable]

abstract class KAddr
case object Halt extends KAddr
case class ContAddr(tgt: Expr, time: Time) extends KAddr

case class Frame(x: String, e: Expr, env: Env)
case class Cont(frame: Frame, kaddr: KAddr)
type KStore = Store[KAddr, Cont]

case class State(e: Expr, env: Env, bstore: BStore, 
                 kstore: KStore, k: KAddr, time: Time)
\end{verbatim}

\begin{verbatim}
def allocBind(x: String, time: Time): BAddr = BAddr(x, time)
def allocKont(tgtExpr: Expr, tgtEnv: Env, tgtStore: BStore, time: Time): KAddr = 
  ContAddr(tgtExpr, time)
def k: Int = 0
def tick(s: State): Time = (s.e::s.time).take(k)

def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())

def atomicEval(e: Expr, env: Env, bstore: BStore): Set[Storable] = e match {
  case Var(x) => bstore(env(x)).toSet
  case lam@Lam(x, body) => Set(Clos(lam, env))
}
\end{verbatim}

\begin{verbatim}
def step(s: State): List[State] = {
  val newTime = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) if isAtomic(ae)=>
      for (Clos(Lam(v, body), c_env) <- atomicEval(f, env, bstore).toList) yield {
        val baddr = allocBind(v, newTime)
        val newEnv = c_env + (v -> baddr)
        val newBStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val newKAddr = allocKont(body, c_env, newBStore, newTime)                                                          
        val newKStore = kstore.update(newKAddr, Cont(Frame(x, e, env), kaddr))
        State(body, newEnv, newBStore, newKStore, newKAddr, newTime)
      }
    case State(ae, env, bstore, kstore, kaddr, time) if isAtomic(ae)=>
      for (Cont(Frame(x, e, f_env), f_kaddr) <- kstore(kaddr).toList) yield {
        val baddr = allocBind(x, newTime)
        val newEnv = f_env + (x -> baddr)
        val newStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, newEnv, newStore, kstore, f_kaddr, newTime)
      }
  }
}
\end{verbatim}

Comparing with concrete CESK machine, the
\texttt{drive} function performs a collecting semantics rather than valuation
semantics. That is, for the purpose of analyzing a program, \texttt{drive}
collects all the intermediate machine states as program is abstractly executing.
This is a variant of work list algorithm that always applies \texttt{step}
function to the first element of \texttt{todo} and adds it to \texttt{seen}
if there is an unexplored one. If the work list is empty, the function just
returns the set of explored states up to now.

\begin{verbatim}
def drive(todo: List[State], seen: Set[State]): Set[State] = todo match {
  case Nil => seen
  case hd::tl if seen.contains(hd) => drive(tl, seen)
  case hd::tl => drive(step(hd).toList ++ tl, seen + hd)
}

def analyze(e: Expr): Set[State] = drive(List(inject(e)), Set())
\end{verbatim}

%\subsection{Monadic Abstract Interpreter}
%Maybe not here?

@@

describe problem, show some examples, why it is important

In this section, we firstly describe a naive and uncomputable solution to the
problem; then briefly review several recent research works that enable
computable and precise call/return match. Even though the first solution is
uncomputable, but as we will see in section 3, it is the starting point of our
transformation.

@@

uncomputable solutions

\begin{verbatim}
case class State(e: Expr, env: Env, bstore: BStore, konts: List[Frame], time: Time)

def step(s: State): List[State] = {
  val newTime = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) =>
      for (Clos(Lam(v, body), c_env) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, newTime)
        val newEnv = c_env + (v -> baddr)
        val newStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, newEnv, newStore, frame::konts, newTime)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, f_env)::konts =>
          val baddr = allocBind(x, newTime)
          val newEnv = f_env + (x -> baddr)
          val newStore = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, newEnv, newStore, konts, newTime))
      }
  }
}

\end{verbatim}

\subsubsection{Computable Solutions}

Recent years, there are great efforts \cite{vardoulakis2010cfa2,
  earl2012introspective, gilray2016pushdown}
to achieve precise call/return match based
on small-step abstracting abstract machines which is we described in section \ref{aam}.

go over existing computable solutions P4F ANF, not works on direct style,PDCFA,AAC

@@

specify higher-order definitional interpreters using first-order means (cite)
